---
title: "ENVS 193DS Spring 2025 Final"
author: Ellie Schomberg
date: 06-11-2025
format: 
  html:
    toc: true
---

Link to the GitHub repository is [here](https://github.com/ellieschomberg/ENVS-193DS_spring-2025_final)

# Set Up

```{r set up}
#| message: FALSE
# reading in packages
library(tidyverse)
library(janitor)
library(here)
library(flextable)
library(readxl)
library(DHARMa)
library(MuMIn)
library(ggeffects)

# reading in data
sst <- read.csv(here("data", "SST_update2023.csv"))
nest_boxes <- read.csv(here("data", "occdist.csv"))
```

# Problem 1. Research Writing

## a. Transparent statistical methods

In part 1, they used a Pearson's r test. In part 2, they used an ANOVA.

## b. More information needed

One additional piece of information that should be included in the results section is a post hoc test, like Tukey's HSD. We know from the results of the ANOVA that at least two of the groups are different from each other, but not which ones. The Tukey's HSD would provide information as to which groups are significantly different from each other. In this example, a Tukey's HSD would be able to tell you which two sources between urban land, atmospheric deposition, fertilizer, wastewater treatment, and grassland were significantly different from each other in terms of average nitrogen load. Another additional piece of information that could be included in the results is the effect size, which quantifies how much the grouping variable affects the response variable. In this example, the effect size would tell whether the different sources had a negligible, small, medium, or large effect on the nitrogen load.

## c. Suggestions for rewriting

Part 1:

There is a [weak, moderate, strong] [positive, negative] relationship between the distance from headwater (km) and the annual total nitrogen load (kg/year) (Pearson's r = correlation coefficient, p = 0.03, $\alpha$ = significance level).

Part 2:

We found a [small, medium, large] difference ($\eta^2$ = effect size) between sources in average nitrogen load (one way ANOVA, F(num df, denom df) = F-statistic, p = 0.02, $\alpha$ = significant level). On average (one of the significantly different sources) tended to have a [higher or lower] average nitrogen load than (other significantly different source) (ex results from Tukey's HSD (difference) kg/year, 95% CI: [value, value] kg/year).


# Problem 2. Data visualization

## a. Cleaning and summarizing

```{r cleaning and summarizing}
#| message: FALSE
sst_clean <- sst |>  # starting with sst data frame
  mutate(
    month = month(date, label = TRUE, abbr = TRUE) |>  # adding a column for months
              factor(levels = month.abb, ordered = TRUE), # creating the months as an ordinal factor with name abbreviations
    year = year(date)) |> # creating a column for years
  group_by(month, year) |>  # grouping the data by month and year
  summarize(mean_monthly_sst = mean(temp, na.rm = TRUE))  |> # creating the mean monthly sst column
  ungroup() |> 
  filter(year %in% c("2018", "2019", "2020", "2021", "2022", "2023")) |> # filtering to include 2018-2023
  mutate( year = as_factor(year), # making the years a factor
          year = fct_relevel(year,"2018", "2019", "2020", "2021", "2022", "2023")) |> 
  select( year, month, mean_monthly_sst) # selecting the columns of interest

slice_sample(sst_clean, n=5) 
str(sst_clean)
```


## b. Visualize the data

```{r visualization}
#| warning: FALSE
ggplot(data = sst_clean, # starting with sst_clean data frame
       aes(x= month, # x axis is month
           y = mean_monthly_sst, # y axis is mean monthly sst
           color = year, # color and group data by year
           group = year)) +
  geom_point() + # adding points
  geom_line() + # addding lines
  scale_color_manual(values = c("2023" = "steelblue4", # creating color gradient
                                "2022" = "steelblue", 
                                "2021" = "steelblue3", 
                                "2020" = "steelblue2",  
                                "2019" = "steelblue1", 
                                "2018" = "lightskyblue")) +
  labs(x = "Month", # renaming x axis
       y = "Mean monthly sea surface temperature (°C)", # renaming y axis
       color = "Year") + # renaming legend
  theme(panel.background = element_blank(), # making background blank
        panel.border = element_rect(color = "gray", fill = NA, linewidth = 1), # adding border
        axis.line = element_blank(), # getting rid of gridlines
        legend.position = c(0.05, 0.95), # moving legend to the top left inside the panel
        legend.justification = c(0, 1))
```


# Problem 3. Data analysis

```{r cleaning and wrangling}
#| echo: false
nest_boxes_clean <- nest_boxes |> # creating new object starting with nest_boxes data frame
  clean_names() |> # cleaning column names
  mutate(box_occupant = as_factor(box_occupant), # making the box occupant  a factor
         box_occupant = fct_relevel(box_occupant,
                                    "swift parrot", 
                                    "common starling",
                                    "tree martin",
                                    "empty")) |> 
  mutate(season = as_factor(season), # making the season a factor
         season = fct_relevel(season,
                              "2016",
                              "2019")) |> 
  select(box_occupant, edge_distance, season, sp, cs, e, tm) # selecting the columns of interest
```

```{r visualizing data}
#| echo: false
#| output: false
# what is the relationship between distance and occupancy for each species
ggplot(data = nest_boxes_clean, # data is nest boxes clean
       aes(x = box_occupant, # x axis is box occupant
           y = edge_distance, # y axis is edge distance
           color = box_occupant)) + # grouping by the different species
  geom_jitter(width = 0.1, # adding points
              height = 0,
              shape = 21,
              alpha = 0.6) +
  stat_summary(geom = "point", # adding means
               fun = mean,
               size = 4)
```


## a. Response variable

In this data set, 0s and 1s represent absence and presence, respectively, of the different bird species in the nest boxes. For example, if a Tree Martin was present, there would be a 1 in that column, and if there were none present, there would be a 0 in that column.

## b. Purpose of study

In the context of this study, Swift Parrots are the target species, as they are on the brink of extinction, and the nest boxes are going to be used in theory to help their population recover. The other two species, the Common Starling and the Tree Martin, are potential competitors that may interfere with the nest boxes.

## c. Difference in "seasons"

The two different "seasons" that were compared in this study were the years 2016 and 2019. 2016 was the year that the nest boxes were initially deployed, and 2019 was 3 years after. In 2019, a tree flowering event triggered a second Swift Parrot breeding event at the study site.

## d. Table of models

| Model Number | Season | Edge Distance | Predictor List               |
|:------------:|:------:|:-------------:|:----------------------------:|
| 0            |        |               | no predictors (null model)   |
| 1            |   X    |       X       | all predictors (full model)  |
| 2            |   X    |               | Season                       |
| 3            |        |       X       | Edge Distance                |


## e. Run the models

```{r models}
#model 0: null model 
model0 <- glm(sp ~ 1, # formula
              data = nest_boxes_clean, # data set
              family = "binomial") # binomial data

# model 1: all predictors
model1 <- glm(sp ~ edge_distance + season, # formula
              data = nest_boxes_clean, # data set
              family = "binomial") # binomial data

# model 2: season
model2 <- glm(sp ~ season, # formula
              data = nest_boxes_clean, # data set
              family = "binomial") # binomial data

# model 3: edge distance
model3 <- glm(sp ~ edge_distance, # formula
              data = nest_boxes_clean, # data set
              family = "binomial") # binomial data
```

## f. Check the diagnostics

```{r diagnostics}
#| message: false
# diagnostics for model 0
plot(
  simulateResiduals(model0))

# diagnostics for model 1
plot(
  simulateResiduals(model1))

# diagnostics for model 2
plot(
  simulateResiduals(model2))

#diagnostics for model 3
plot(
  simulateResiduals(model3))
```


## g. Select the best model

```{r model selection}
AICc(model0,
     model1, # best model: all predictors
     model2,
     model3) |> 
  arrange(AICc) # arranging output


```

```{r}
#| echo: false
#| output: false
gtsummary::tbl_regression(model1,
                          exponentiate = TRUE)
summary(model1)
```

The best model as determined by Akaike's Information Criterion (AIC) is the model that includes both of the predictor variables, season and distance from the forest edge, to predict whether or not a Swift Parrot was present in the nest box.

## h. Visualize the model predictions

```{r model visualization}
#| message: FALSE
model1_predictions <- ggpredict(model1, # creating object for model 1 predictions
          terms = c("edge_distance", "season")) # predictor variables

ggplot(data = nest_boxes_clean, # start with next boxes clean data frame
       aes(x = edge_distance, # x axis is edge distance
           y = sp)) + # y axis is swift parrot presence/absence
  geom_point(aes(color = season), # adding points
             size = 3,
             alpha = 0.4) +
  geom_ribbon(data = model1_predictions, # adding ribbon
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = group),
                  alpha = 0.4) +
  geom_line(data = model1_predictions, # adding line
            aes(x = x,
                y = predicted,
                color = group)) +
  scale_y_continuous(limits = c(0,1), # setting limits and only having 2 tick marks
                     breaks = c(0, 1)) +
  scale_color_manual(values = c("2016" = "mediumpurple", "2019" = "royalblue")) +
  scale_fill_manual(values = c("2016" = "mediumpurple1", "2019" = "royalblue1")) +
  labs(x = "Distance from the forest edge (m)", # renaming title, axis, legend
       y = "Probability of box occupancy", 
       title = "Swift Parrot",
       color = "season",
       fill = "season") +
  theme(panel.background = element_blank(), # getting rid of background
        axis.line = element_blank(), # getting rid of grid lines and ticks
        axis.ticks = element_blank())

```


## i. Write a caption for your figure

**Figure 1. Probability of Swift Parrot nest box occupancy is higher in 2016 than 2019 and at closer distances to the forest edge.**
Data from Stojanovic, Dejan et al. (2021). Do nest boxes breed the target species or its competitors? A case study of a critically endangered bird [Dataset]. Dryad. https://doi.org/10.5061/dryad.83bk3j9sb. Points represent individual observations of the presence or absence of a swift parrot in a nesting box. Lines represent the model predictions for Swift Parrot occupation. Ribbons represent the 95% confidence interval. Colors represent the season (Purple = 2016, Blue = 2019).

## j. Calculate model predictions

```{r model predictions}
# probability of occupancy at 0m in 2016
ggpredict(model1,
          terms = c("edge_distance [0]", "season [2016]"))
# probability of occupancy at 900m in 2016
ggpredict(model1,
          terms = c("edge_distance [900]", "season [2016]"))
# probability of occupancy at 0m in 2019
ggpredict(model1,
          terms = c("edge_distance [0]", "season [2019]"))
# probability of occupancy at 900m in 2019
ggpredict(model1,
          terms = c("edge_distance [900]", "season [2019]"))
```


## k. Interpret your results

Swift parrots tend to occupy nest boxes shortly after their deployment in their nesting sites, and at closer distances to the forest edge (Figure 1).

In the 2016 season, the probability of a swift parrot occupying a nest box 0m from the forest edge is 0.48 (95% CI :[0.33, 0.64]]) and 900m from the forest edge is 0.12 (95% CI: [0.06, 0.24]). In the 2019 season, the probability of a swift parrot occupying a nest box 0m from the forest edge is 0.30 (95% CI: [0.18, 0.44]) and 900m from the forest edge is 0.06 (95% CI: [0.03, 0.13]).

The occupancy trends of the Swift Parrot can be partly explained by the behavior of its common competitors. One of the Swift Parrot's competitors, the Tree Martin, tends to occupy the nest boxes at further distances away from the forest edge. In addition, their other competitor, the Common Starling, tend to occupy the older nest boxes. Because of this, Swift Parrots tend to occupy the newer nesting boxes that are closer to the edge of the forest.


# Problem 4. Affective and explanatory visualizations

## a. Comparing visualizations

In Homework 2, I represented my data as a bar graph and a line graph, because I did not have many observations to go off of at that point. By the time I made my affective visualization in Homework 3, I had finished all of my data collection. My affective visualization is a paiting that looks like a swimming pool, but my underlying data is represented in a jitterplot.

Some of the similarities that I see between my exploratory visualizations and my affective visualization is that they show the comparison between categories. Both the bar graph and the jitterplot in my affective visualization were used to compare the mean amount of sleep in hours between the different categories of swimming workouts that I completed.

Even though I did not have many observations in my initial exploratory visualization, I could see from the bar graph that the Aerobic workout lead to the highest average hours slept, and from my line graph, an increased distance in yards also lead to the highest amount of hours slept. I noticed similar patterns in my final affective visualization, where the Aerobic and Racing categories of swimming workouts had the highest means, both with an average of 7.8 hours. All of my data matched my predictions, that longer, more intense workouts lead me to sleep a greater amount of hours at night.

One piece of feedback that I got during week 9 workshop was to include the individual data observations in my visualization. Before, I had just had a swimmer that I used to represent the mean in each category, by showing them at different places within a lane that represented a different category. I decided to add little bubbles to each of the lanes, which represent the individual data points, and I think adding this component really helped to make the visualization more clear.

